\\fontsize{12pt}{3pt}\\selectfont \\textbf{\\textit{Submitted to Ecosphere. Manuscript type: Article}}

### 2023-09-24 manuscript/manuscriptRW.Rmd
changes in both reviewers.

Despite considerable mismatches between simulation results and data, it
is remarkable that most models capture the essential features of the
highly complex process of tree regeneration, while not having been
parameterized with such data. We conclude that much can be gained by
evaluating and refining the modeling of regeneration processes. This has
the potential to render long-term projections of forest dynamics under
changing environmental conditions that are much more robust.

@@

Thus, a focus on the modeling of tree regeneration processes is sorely
needed and overdue [@price2001; @walck2011] if we are to make reliable
projections of future forest dynamics, i.e., when the models need to be
operated in extrapolation mode, as well as from a fundamental ecological
point of view for increasing systems understanding. Also in empirical
ecology, tree regeneration has not received a lot of attention
[@hanbury-brown2022], although many sub-processes have been studied in
detail [@miina2006]. Yet, little data are available that covers all the
processes within one species along environmental gradients, let alone
for a vast suite of species. Thus, the available data are fragmented,
which constitutes a major problem for model building [@clark1999]. On
the one hand, monitoring on permanent plots (such as National Forest
Inventories) often measures tree regeneration [@zell2019], however with
a design that captures the rate after the stand initiation phase
[@hallsby2015]. Data from permanent plots, although available across
regions, are highly heterogeneous, rendering their use difficult in a
modeling context [@käber2021]. On the other hand, targeted studies to
measure tree regeneration on experimental sites are often limited in
spatial extent, species studied, or the sub-set of processes that are
targeted [@berdanier2016; @collet2006].

In the present study, for the first time, a large number of forest
models commonly used to assess forest dynamics under climate change are
evaluated against a continental-scale, multi-species harmonized dataset
on tree regeneration [@käber2023]. By tree regeneration, we refer to the
passing of trees across a specific diameter threshold ("ingrowth"). We
included models that are based on a range of "philosophies", from models
operating at the stand to the global scale as well as the range of
models from empirically derived to "process-based" [e.g., @bugmann1996;
@fabrika2012; @lexer2001; @reyer2014]. Due to the large variability in
tree regeneration patterns in nature and the large number of factors
driving and constraining this process --- including some that are not
incorporated explicitly in most models, such as deer browsing --- we do
not aim for a detailed statistical evaluation of each model. Instead, we
aim to evaluate the general regeneration patterns and magnitudes
simulated by the models, and benchmark the simulated regeneration niche
of multiple species against empirical data along a wide environmental
gradient of temperature, moisture, and light availability [@grubb1977].

More specifically, we aim to answer the following questions: (1) Are
models of forest dynamics capturing accurately tree regeneration levels,
initial tree species diversity and mortality in the regeneration? (2) Do
model traits explain differences in model performance? (3) How well do
the models capture total regeneration and individual species
regeneration niches across environmental gradients of light
availability, temperature, and soil moisture?

@@

We selected 200 sites from this network as the benchmarking dataset for
the simulation to be representative of the environmental variation
contained in the data. This was achieved by applying k-means clustering
to define 200 clusters of plots from the original set of 869 plots along
the environmental dimensions of temperature, climatic water balance,
soil quality, slope and aspect. Each of these sites featured at least
two consecutive measurements. Regeneration minimum size thresholds for
these sites differed between stem diameters of 0 and 10 cm. For our
study, we defined two datasets where one included 165 sites with a
diameter threshold of 7 cm or lower, and the other included another 35
sites with diameter thresholds between 7 and 10 cm.

@@

The simulations were run for mixed-species forests (not multiple
single-species simulations) using mixtures of eleven species or genera
for which regeneration data of sufficient quality were available from
EuFoRIa: *Fagus sylvatica L., Picea abies L., Abies alba Mill., Carpinus
betulus L., Tilia cordata Mill., Acer pseudoplatanus L., Betula spp. L.,
Fraxinus excelsior L., Quercus spp. L., Alnus glutinosa L., Pinus
sylvestris L.* The same set of species was used at all 200 sites. Two
models included their standard set of species for the simulations, which
is much larger (i.e., ForClim 1, ForClim 11, TreeMig). In three models,
fewer than these eleven species were simulated (4C, xComp, LPJ-GUESS)
(Table \@ref(tab:tableModels)). In 4C only *Fagus sylvatica*, *Picea
abies*, *Betula* spp., *Quercus* spp. and *Pinus sylvestris* are
parameterised. xComp simulations did not consider *Fraxinus excelsior*
due to a depreciated species parameterisation. In LPJ-GUESS, *Acer
pseudoplatanus* and *Alnus glutinosa* are not parameterized and
therefore these species could not be included in the simulation.
Finally, the model aDGVM2 does not simulate individual species. Rather,
community assembly processes and trait filtering generate plant
communities that are adjusted to the biotic and abiotic conditions, and
the simulated plants can be classified into ecological strategies based
on their trait values in a post-processing step.

@@

```{=tex}
\begin{equation}
\bar{R}_{BA_{share_{t,i}}} = \frac{\bar{r}_{BA_{t,i}}}{\sum_{i = 1}^{s}{\bar{r}_{{BA_{t,i}}}}} \label{eq:BAshare}
\end{equation}
```
where $s$ is the total number of species simulated, and $\bar{r}_{t,i}$
is the mean basal area in the regeneration (subsequently referred to as
"regeneration basal area") of species $i$ per site $t$ across the
available samples at that site. Furthermore, regarding the share in the
regeneration basal area per species ($\bar{R}_{BA_{share_{i}}}$), we
categorized this as zero when the regeneration basal area of that
species ($\bar{r}_{BA_{t,i}}$) was zero, but also when both the
regeneration basal area of that species ($\bar{r}_{BA_{t,i}}$) and the
total regeneration basal area ( $\sum_{i = 1}^{s}{\bar{r}_{BA_{t,i}}}$)
equaled zero.

@@

The observed regeneration did not change much across the climatic water
balance gradient, and showed a slight increase with temperature. The
models did not match the observed regeneration patterns across the
gradients of soil moisture and temperature (Figure
\@ref(fig:nvTrend710BAlightTemp)b,c and S7), and neither was there a
clear pattern across models. Regarding the soil moisture gradient, a
group of models featured decreasing regeneration with increasing soil
moisture (4C, ForClim 1, ForClim11, SIBYLA, xComp, PICUS, LandClim,
TreeMig and iLand), while a few models showed the opposite trend
(ForCEEPS(f), FORMIND and LPJ-GUESS). Across the temperature gradient,
all the models that featured a decrease in regeneration levels with
increasing soil moisture showed the opposite trend (i.e., more
regeneration with increasing temperature), except for 4C. The other
models did not feature a clear pattern.

@@

In spite of the critical considerations above, it is remarkable that
most models did not deviate exceedingly from the observations with
respect to simulated regeneration levels -- after all, few if any of the
models' regeneration routines are sufficiently constrained by data, with
the exception of the empirical models xComp and SIBYLA. It is noteworthy
that in spite of their empirical basis, these two models did not match
empirical regeneration levels, in a similar magnitude as the other
models and even in opposite directions (SIBYLA: underestimation; xComp:
overestimation). It appears that using empirical data of limited
geographical scope and constrained to managed forests, as done in these
models, leads to extrapolation problems already under current climatic
conditions. For the other models, whose regeneration modules are not
strongly constrained by empirical data, the degrees of freedom that are
available for modeling regeneration [@konig_tree_2022] could have
implied that model performance would be much worse than what we found.

@@

There were pronounced differences how the main tree species were
represented by the models along the environmental gradients, in
particular the dominant species *Abies alba*, *Fagus sylvatica* and
*Picea abies*. The regeneration levels were sampled from simulations in
the equilibrium, and in this state it is expected that non-climax
species such as *Pinus sylvestris* or *Quercus* would be of minor
importance, or absent [@klopcic2015]. Most models captured this low
abundance, which is also found in the empirical data. Thus, although the
broad patterns are matched by many models, improvements in the
quantification of the regeneration niche of the species are needed, but
this cannot be done in the absence of robust datasets across multiple

@@

#footnotes
names(modelDF)[2] <- paste0(names(modelDF)[2], "(a)")
names(modelDF)[3] <- paste0(names(modelDF)[3], "(b)")
names(modelDF)[4] <- paste0(names(modelDF)[4], "(c)")
names(modelDF)[6] <- paste0(names(modelDF)[6], "(d)")
names(modelDF)[11] <- paste0(names(modelDF)[11],"(e)")

kable(modelDF,
      format = "latex", booktabs = TRUE,
      caption = "Models included and their characteristics.") |>
        kableExtra::landscape() |>
        kableExtra::kable_styling(full_width = F, latex_options = "scale_down") |>
    # fix width by column
      kableExtra::column_spec(column = 1, width = "5em") |>
      kableExtra::column_spec(column = 2, width = "3em") |>
      kableExtra::column_spec(column = 3, width = "4em") |>
      kableExtra::column_spec(column = 4, width = "5em") |>
      kableExtra::column_spec(column = 5, width = "5em") |>
      kableExtra::column_spec(column = 6, width = "5em") |>
      kableExtra::column_spec(column = 7, width = "8em") |>
      kableExtra::column_spec(column = 8, width = "8em") |>
      kableExtra::column_spec(column = 9, width = "5em") |>
      kableExtra::column_spec(column = 10, width = "5em") |>
      kableExtra::column_spec(column = 11, width = "10em") |>
      kableExtra::add_footnote(c("S-Stand, L-Landscape, G-Global",
                                 "PB-Process based, EM-Empirical",
                                 "C-Cohort, I-Individual",
                                 "Y-Yes, N-No",
                                 "D-Daily, M-Monthly"),

@@

# Figure captions

@@

**Figure 4.** Mean Shannon index across all samples per site for
observed and simulated data. Each plot shows one pattern represented by
one exemplary model of each category (overperdicted, intermediate and
underpredicted). The full data with the grouping of the models are shown
in Figures S3 and S4. n indicates the number of models falling in each
group. A) Examples for the three trends across models for species
diversity at the stand level. B) Examples for the three trends across
models for regeneration at the 7 cm threshold.

@@

**Figure 7.** Mean regeneration values across the 200 samples per site
with the y-axis scaled differently by model, for the 200 sites against
gradients of (a) total basal area; (b) climatic water balance; (c)
seasonal degree-day sum. The values were split into ten bins; the red
lines represent a GAM model showing the trend in the observed data. For
comparing the magnitude of recruitment across models the same graphs
with equally scaled y-axis can be found in Figure S7.

>>>>

Despite the mismatches between simulation results and data, it is remarkable that most models capture the essential features of the highly complex process of tree regeneration, while not having been parameterized with such data. We conclude that much can be gained by evaluating and refining the modeling of regeneration processes. This has the potential to render long-term projections of forest dynamics under changing environmental conditions that are much more robust.

@@

Modeling tree regeneration processes is challenging, and even in empirical ecology, it has not received much attention [@hanbury-brown2022], although many sub-processes have been studied in detail [@miina2006]. Yet, little data are available that covers all the processes within one species along environmental gradients, let alone for a vast suite of species. Data on forest regeneration data are often fragmented, which constitutes a major problem for model building [@clark1999a]. For example, monitoring on permanent plots (such as National Forest Inventories) often measures tree regeneration [@zell2019], however with a design that captures the rate after the stand initiation phase [@hallsby2015]. Data from permanent plots, although available across regions, are also highly heterogeneous, rendering their use difficult in a modeling context [@käber2021], and targeted studies to measure tree regeneration on experimental sites are often limited in spatial extent, species studied, or the sub-set of processes that are targeted [@berdanier2016; @collet2006].

Thus, a focus on the modeling of tree regeneration processes is sorely needed and overdue [@price2001; @walck2011] if we are to make reliable projections of future forest dynamics, i.e., when the models need to be operated in extrapolation mode, as well as from a fundamental ecological point of view for increasing systems understanding. In the present study, for the first time, a large number of forest models commonly used to assess forest dynamics under climate change are evaluated against a continental-scale, multi-species harmonized dataset on tree regeneration [@käber2023]. By tree regeneration, we refer to the passing of trees across a specific diameter threshold ("ingrowth"). We included models that are based on a range of "philosophies", from models operating at the stand to the global scale as well as the range of models from empirically derived to "process-based" [e.g., @bugmann1996; @fabrika2012; @lexer2001; @reyer2014]. Due to the large variability in tree regeneration patterns in nature and the large number of factors driving and constraining this process --- including some that are not incorporated explicitly in most models, such as deer browsing --- we do not aim for a detailed statistical evaluation of each model. Instead, we aim to evaluate the general regeneration patterns and magnitudes simulated by the models, and benchmark the simulated regeneration niche of multiple species against empirical data along a wide environmental gradient of temperature, moisture, and light availability [@grubb1977].

More specifically, we aim to answer the following questions: (1) Are models of forest dynamics capturing accurately tree regeneration levels, initial tree species diversity and mortality in the regeneration? (2) Do model traits explain differences in model performance? (3) How well do the models capture total regeneration and individual species regeneration niches across environmental gradients of light availability, temperature, and soil moisture? We evaluate the performance of regeneration models by analyzing simulated data alongside observed data. Specifically, we examine tree species diversity, mortality rates in tree regeneration, and regeneration levels. Then, we link the model's traits and performance to define regeneration and we assess the regeneration gradients on regeneration niches. Finally, we provide research recommendations for next steps towards improving tree regeneration modules.

@@

We selected 200 sites from this network as the benchmarking dataset for the simulation to be representative of the environmental variation contained in the data. This was achieved by applying k-means clustering to define 200 clusters of plots from the original set of 869 plots along the environmental dimensions of temperature, climatic water balance, soil quality, slope and aspect. Each of these sites featured at least two consecutive measurements. Regeneration minimum size thresholds for these sites differed between stem diameters of 0 and 10 cm. For our study, we defined two datasets where one included 165 sites with a diameter threshold of 7 cm or lower, and the other included another 35 sites with diameter thresholds from 10 cm. Therefore we had in total 200 sites with observations of newly established trees above 10 cm and 165 sites above 7cm. In the results section any result from observed data at 7 cm threshold is originated from the 165 sites.

@@

The simulations were run for mixed-species forests (not multiple single-species simulations) using mixtures of eleven species or genera for which regeneration data of sufficient quality were available from EuFoRIa: *Fagus sylvatica L., Picea abies L., Abies alba Mill., Carpinus betulus L., Tilia cordata Mill., Acer pseudoplatanus L., Betula spp. L., Fraxinus excelsior L., Quercus spp. L., Alnus glutinosa L., Pinus sylvestris L.* The same set of species was used at all 200 sites. Two models included their standard set of species for the simulations, which is much larger (i.e., ForClim 1, ForClim 11, TreeMig), the species simulated beyond the eleven species were lumped in a category "others". In three models, fewer than these eleven species were simulated (4C, xComp, LPJ-GUESS) (Table \@ref(tab:tableModels)). In 4C only *Fagus sylvatica*, *Picea abies*, *Betula* spp., *Quercus* spp. and *Pinus sylvestris* are parameterised. xComp simulations did not consider *Fraxinus excelsior* due to a depreciated species parameterisation. In LPJ-GUESS, *Acer pseudoplatanus* and *Alnus glutinosa* are not parameterized and therefore these species could not be included in the simulation. Finally, the model aDGVM2 does not simulate individual species. Rather, community assembly processes and trait filtering generate plant communities that are adjusted to the biotic and abiotic conditions, and the simulated plants can be classified into ecological strategies based on their trait values in a post-processing step.

@@

```{=tex}
\begin{equation}
\bar{R}_{BA_{share_{t,i}}} = \frac{\bar{r}_{BA_{t,i}}}{\sum_{i = 1}^{s}{\bar{r}_{{BA_{t,i}}}}} \label{eq:BAshare}
\end{equation}
```
where $s$ is the total number of species simulated, and $\bar{r}_{t,i}$ is the mean basal area in the regeneration (subsequently referred to as "regeneration basal area") of species $i$ at site $t$ across the available samples at that site. Furthermore, regarding the share in the regeneration basal area per species ($\bar{R}_{BA_{share_{i}}}$), we categorized this as zero when the regeneration basal area of that species ($\bar{r}_{BA_{t,i}}$) was zero, but also when both the total regeneration basal area ( $\sum_{i = 1}^{s}{\bar{r}_{BA_{t,i}}}$) and the regeneration basal area of that species ($\bar{r}_{BA_{t,i}}$) equaled zero.

@@

The observed regeneration did not change much across the climatic water balance gradient, and showed a slight increase with temperature. The models did not match the observed regeneration patterns across the gradients of soil moisture and temperature (Figure \@ref(fig:nvTrend710BAlightTemp)b,c), and neither was there a clear pattern across models. Regarding the soil moisture gradient, a group of models featured decreasing regeneration with increasing soil moisture (4C, ForClim 1, ForClim11, SIBYLA, xComp, PICUS, LandClim, TreeMig and iLand), while a few models showed the opposite trend (ForCEEPS(f), FORMIND and LPJ-GUESS). Across the temperature gradient, all the models that featured a decrease in regeneration levels with increasing soil moisture showed the opposite trend (i.e., more regeneration with increasing temperature), except for 4C. The other models did not feature a clear pattern.

@@

In spite of the critical considerations above, it is remarkable that most models did not deviate exceedingly from the observations with respect to simulated regeneration levels -- after all, few if any of the models' regeneration routines are sufficiently constrained by data, with the exception of the empirical models xComp and SIBYLA. It is noteworthy that in spite of their empirical basis, these two models did not match empirical regeneration levels, in a similar magnitude as the other models and even in opposite directions (SIBYLA: underestimation; xComp: overestimation). It appears that using empirical data of limited geographical scope and constrained to managed forests, as done in these models, leads to extrapolation problems already under current climatic conditions. For the other models, whose regeneration modules are not strongly constrained by empirical data, the multiple strategies that are available for modeling regeneration [@konig_tree_2022] could have implied that model performance would be much worse than what we found.

@@

There were pronounced differences how the main tree species were represented by the models along the environmental gradients, in particular the dominant species *Abies alba*, *Fagus sylvatica* and *Picea abies*. The regeneration levels were sampled from simulations in the equilibrium, and in this state it is expected that non-climax species such as *Pinus sylvestris* or *Quercus* would be of minor importance, or absent [@klopcic2015]. Most models captured this low abundance, which is also found in the empirical data. Thus, although the broad patterns are matched by many models, improvements in the quantification of the regeneration niche of the species are needed, but this cannot be done in the absence of robust datasets across multiple species.

@@

#footnotes
names(modelDF)[2] <- paste0(names(modelDF)[2], "(a)")
names(modelDF)[3] <- paste0(names(modelDF)[3], "(b)")
names(modelDF)[4] <- paste0(names(modelDF)[4], "(c)")
names(modelDF)[5] <- paste0(names(modelDF)[5], "(d)")
names(modelDF)[6] <- paste0(names(modelDF)[6], "(e)")
names(modelDF)[11] <- paste0(names(modelDF)[11],"(f)")

kable(modelDF,
      format = "latex", booktabs = TRUE,
      caption = "") |>
        kableExtra::landscape() |>
        kableExtra::kable_styling(full_width = F, latex_options = "scale_down") |>
    # fix width by column
      kableExtra::column_spec(column = 1, width = "5em") |>
      kableExtra::column_spec(column = 2, width = "3em") |>
      kableExtra::column_spec(column = 3, width = "4em") |>
      kableExtra::column_spec(column = 4, width = "5em") |>
      kableExtra::column_spec(column = 5, width = "5em") |>
      kableExtra::column_spec(column = 6, width = "6em") |>
      kableExtra::column_spec(column = 7, width = "5em") |>
      kableExtra::column_spec(column = 8, width = "8em") |>
      kableExtra::column_spec(column = 9, width = "8em") |>
      kableExtra::column_spec(column = 10, width = "5em") |>
      kableExtra::column_spec(column = 11, width = "5em") |>
      kableExtra::column_spec(column = 12, width = "10em") |>
      kableExtra::add_footnote(c("S-Stand, L-Landscape, G-Global",
                                 "PB-Process based, EM-Empirical",
                                 "C-Cohort, I-Individual",
                                 "Y-Yes, N-No",
                                 "R-Regeneration, Rc-Recruitment",
                                 "D-Daily, M-Monthly"),

@@

# Table and Figure captions

**Table 1.** Models included, their characteristics and simulation strategies used in this study. Where Scale refers to the models scale, Type to the models type, Population structure to the model population approach, Feedback to the inclusion or not of adult tree feedback to the level and species composition of regeneration, Approach to the model regeneration module approach, Start from to the stage from which trees are recruited, Complexity, to the mean regeneration formulation as estimated in (Bugmann & Seidl, 2022), Species to which species from the original eleven species were simulated , Runtime (spin up) [years] to the number of years used per model in their simulations for the spin up period and regeneration sampling; Climate data type (use of years) to the type of data used in the simulations and how it was used, and Reference provides the link to the publication with more model details. Further information can be found on the individual model reports available in the supplementary materials.

@@

**Figure 4.** Mean Shannon index across all samples per site for observed and simulated data. Each plot shows one pattern represented by one exemplary model of each category (overpredicted, intermediate and underpredicted). The full data with the grouping of the models are shown in Figures S3 and S4. n indicates the number of models falling in each group. A) Examples for the three trends across models for species diversity at the stand level. B) Examples for the three trends across models for regeneration at the 7 cm threshold (165 sites).

@@

**Figure 7.** Mean regeneration values across the 200 samples per site with the y-axis scaled differently by model, for the 200 sites against gradients of (a) total basal area; (b) climatic water balance; (c) seasonal degree-day sum. The values were split into ten bins; the red lines represent a GAM model showing the trend in the observed data. Y-axis are not equally scaled.
